Supplementary Materials from: \textit{Innovation and elaboration on the avian tree of life}; Science Advances, in review

### 2023-06-09 Manuscript/second_submission.tex
Create second_submission.tex

\noindent {\normalsize \sc
Thomas Guillerme$^{1,*}$, 
Jen A. Bright$^{2}$,
Christopher R. Cooney${^1}$,
Emma C. Hughes${^1}$,
Zo\"{e} K. Varley${^3}$,
Natalie Cooper$^{3,+}$,
Andrew P. Beckerman$^{1,+}$,
and Gavin H. Thomas$^{1,4,+}$}\\
\noindent {\small \it 
$^1$School of Biosciences, University of Sheffield, Sheffield, S10 2TN, United Kingdom.\\
$^2$School of Natural Science, University of Hull, Hull, HU6 7RX, United Kingdom.\\
$^3$Natural History Museum, Cromwell Road, London, SW7 5BD, United Kingdom.\\
$^4$Bird Group, Department of Life Sciences, the Natural History Museum at Tring, Tring, United Kingdom.\\}

@@

\subsection{Megaevolution: multiple routes to innovation throughout avian evolutionary history}
If deep-time jumps in beak shape \cite{cooney2017mega} are the result of innovation we would expect to see changes in the orientation of trait space of clades (i.e. orders or super-orders) relative to the phylogenetic major axis of beak variation of the whole class Aves.
To test this, we measured the $elaboration_{clade}$ and $innovation_{clade}$ by translating the phylogenetic major axis of beak variation of each clade's (i.e. super-orders and orders) evolutionary rate matrix onto the class-wide phylogenetic major axis of beak variation so that they shared the same origin in the shapespace.
We then used linear algebra to project %@L_linear_algebra_stuff
a focal clade's phylogenetic major axis of beak variation onto the class-wide phylogenetic major axis of beak variation (Fig. S\ref{Fig:cheatsheet}).
We interpret the measured linear algebraic projection (Euclidean distance along the phylogenetic major axis of beak variation) as the clade's $elaboration_{clade}$ score and the measured linear algebraic rejection (Euclidean distance from the phylogenetic major axis of beak variation) as the clade's $innovation_{clade}$ score. % @L_euclidean.
A high $innovation_{clade}$ score indicates that the direction of evolution of a clade differs from the direction of its parent clade (e.g. the direction of evolution of an order differs from its parent super-order).
These scores were calculated for each of 4000 posterior pairs of clade vs.
evolutionary rate matrix (Fig. \ref{Fig:ellipses} boxplots).
We found that for most clades the phylogenetic major axis of beak variation is not aligned with the class-wide phylogenetic major axis of beak variation (Fig. \ref{Fig:ellipses}).
Despite the limitations of representing eight-dimensional space in 2D, our plots of the average elliptical representation of the evolutionary rate matrix for the class-wide phylogenetic major axis of beak variation and each super-order and order (Fig. \ref{Fig:ellipses}b) highlight the striking variation in the orientation of beak shape major axes among clades (Fig. \ref{Fig:orthogonality}).
More than half of the assessed clades (4/8 super-orders and 15/27 orders) displayed higher median $innovation_{clade}$ than median $elaboration_{clade}$ scores implying that $innovation_{clade}$ is a more common generator of beak shape diversity than $elaboration_{clade}$ in deep-time (i.e evidence at the clade scale).

@@

\begin{landscape}
\begin{table}[ht]
\centering
\small
   \begin{tabular}{rrllllllllrr}
     \hline
    Model & $E_{species}$ & $I_{species}$ & $E_{clade}$ & $I_{clade}$ & $E_{species}$ x $I_{clade}$ & $E_{species}$ x $E_{clade}$ & $I_{species}$ x $I_{clade}$ & $I_{species}$ x $E_{clade}$ & $\Delta$AIC & Adj. $r^2$ \\ 
     \hline
   1 & 0.04*** & -0.139*** & 0.002 & -0.002 & 0.064*** & 0.045*** & 0.341*** & 0.093*** & 0.00 & 0.92 \\ 
   2 & 0.257*** & 0.169*** & 0.01 & 0.001 & 0.045*** &  & 0.312*** & 0.034*** & -691.48 & 0.92 \\ 
   3 & 0.108*** & 0.236*** & 0.012 & -0.002 & 0.055*** & 0.034*** & 0.343*** &  & -284.74 & 0.92 \\ 
   4 & 0.254*** & 0.104*** & 0.002 & 0.004 &  & 0.027*** & 0.311*** & 0.052*** & -877.85 & 0.92 \\ 
   5 & 0.1*** & 0.571*** & -0.018* & 0.02*** & 0.054*** & 0.036*** &  & 0.096*** & -1498.36 & 0.91 \\ 
   6 & 0.364*** & 0.973*** & -0.007 & 0.023*** &  &  &  &  & -2230.10 & 0.90 \\ 
   7 & 0.364*** & 0.972*** &  &  &  &  &  &  & -2280.08 & 0.90 \\ 
   8 &  &  & -0.025 & 0.011 &  &  &  &  & -22270.55 & -0.00 \\ 
   9 & 0.286*** &  &  &  &  &  &  &  & -18986.63 & 0.32 \\ 
   10 &  & 0.798*** &  &  &  &  &  &  & -17714.99 & 0.41 \\ 
   11 &  &  & -0.008 &  &  &  &  &  & -22269.34 & -0.00 \\ 
   12 &  &  &  & 0.002 &  &  &  &  & -22269.51 & -0.00 \\ 
      \hline
   \end{tabular}
\caption{Elaboration and innovation as predictors of beak shape distance from centroid.
The table shows parameter estimates of predictors from 12 different phylogenetic generalised least squares (PGLS) models with $\Delta$AIC showing relative model fit following removal of terms. Models 1-5 are multiple regressions including all variables and one or more interaction terms; models 6-8 are are multiple regressions including two or more variables and no interaction terms; models 9-12 are single predictor models. Values in the table are based on PGLS in which Pagel’s $\lambda$ was fixed at 1 (i.e. assuming a Brownian motion model of evolution). Fixing $\lambda$ is necessary to allow model comparison with AIC. We also fitted the same models fixing $\lambda$ to 0.727 (the lowest value of $\lambda$ from any individual model) and relative model fit and interpretation of parameters was unaffected. E = elaboration; I = innovation; AIC = Akaike Information Criterion; NS = not significant; - indicates that the term was not included within that model.}
\label{Tab:pgls}
\end{table}
\end{landscape}

>>>>

\noindent {\normalsize \sc
Thomas Guillerme$^{1,*}$, 
Jen A. Bright$^{2}$,
Christopher R. Cooney${^1}$,
Emma C. Hughes${^1}$,
Zo\"{e} K. Varley${^3,4}$,
Natalie Cooper$^{3,+}$,
Andrew P. Beckerman$^{1,+}$,
and Gavin H. Thomas$^{1,4,+}$}\\
\noindent {\small \it 
$^1$School of Biosciences, University of Sheffield, Sheffield, S10 2TN, United Kingdom.\\
$^2$School of Natural Science, University of Hull, Hull, HU6 7RX, United Kingdom.\\
$^3$Natural History Museum, Cromwell Road, London, SW7 5BD, United Kingdom.\\
$^4$Bird Group, Department of Life Sciences, the Natural History Museum at Tring, Tring, United Kingdom.\\}

@@

\subsection{Megaevolution: multiple routes to innovation throughout avian evolutionary history}
If deep-time jumps in beak shape \cite{cooney2017mega} are the result of innovation we would expect to see changes in the orientation of trait space of clades (i.e. orders or super-orders) relative to the phylogenetic major axis of beak variation of the whole class Aves.
To test this, we measured the $elaboration_{clade}$ and $innovation_{clade}$ by translating the phylogenetic major axis of beak variation of each clade's (i.e. super-orders and orders) evolutionary rate matrix onto the class-wide phylogenetic major axis of beak variation so that they shared the same origin in the shapespace.
We then used linear algebra to project %@L_linear_algebra_stuff
a focal clade's phylogenetic major axis of beak variation onto the class-wide phylogenetic major axis of beak variation (Fig. \ref{Fig:cheatsheet}).
We interpret the measured linear algebraic projection (Euclidean distance along the phylogenetic major axis of beak variation) as the clade's $elaboration_{clade}$ score and the measured linear algebraic rejection (Euclidean distance from the phylogenetic major axis of beak variation) as the clade's $innovation_{clade}$ score. % @L_euclidean.
A high $innovation_{clade}$ score indicates that the direction of evolution of a clade differs from the direction of its parent clade (e.g. the direction of evolution of an order differs from its parent super-order).
These scores were calculated for each of 4000 posterior pairs of clade \textit{vs.} evolutionary rate matrix (Fig. \ref{Fig:ellipses} blue and orange lines).
We found that for most clades the phylogenetic major axis of beak variation is not aligned with the class-wide phylogenetic major axis of beak variation (Fig. \ref{Fig:ellipses}).
Despite the limitations of representing eight-dimensional space in 2D, our plots of the average elliptical representation of the evolutionary rate matrix for the class-wide phylogenetic major axis of beak variation and each super-order and order (Fig. \ref{Fig:ellipses}) highlight the striking variation in the orientation of beak shape major axes among clades (Fig. \ref{Fig:orthogonality}).
More than half of the assessed clades (4/8 super-orders and 15/27 orders) displayed higher median $innovation_{clade}$ than median $elaboration_{clade}$ scores implying that $innovation_{clade}$ is a more common generator of beak shape diversity than $elaboration_{clade}$ in deep-time (i.e evidence at the clade scale).

@@

\begin{landscape}
\begin{table}[ht]
\centering
\footnotesize
   \begin{tabular}{rrllllllllrr}
     \hline
    Model & $E_{species}$ & $I_{species}$ & $E_{clade}$ & $I_{clade}$ & $E_{species}$ $\times$ $I_{clade}$ & $E_{species}$ $\times$ $E_{clade}$ & $I_{species}$ $\times$ $I_{clade}$ & $I_{species}$ $\times$ $E_{clade}$ & $\Delta$AIC & Adj. $r^2$ \\ 
     \hline
   1 & 0.04*** & -0.139*** & 0.002 & -0.002 & 0.064*** & 0.045*** & 0.341*** & 0.093*** & 0.00 & 0.92 \\ 
   2 & 0.257*** & 0.169*** & 0.01 & 0.001 & 0.045*** &  & 0.312*** & 0.034*** & -691.48 & 0.92 \\ 
   3 & 0.108*** & 0.236*** & 0.012 & -0.002 & 0.055*** & 0.034*** & 0.343*** &  & -284.74 & 0.92 \\ 
   4 & 0.254*** & 0.104*** & 0.002 & 0.004 &  & 0.027*** & 0.311*** & 0.052*** & -877.85 & 0.92 \\ 
   5 & 0.1*** & 0.571*** & -0.018* & 0.02*** & 0.054*** & 0.036*** &  & 0.096*** & -1498.36 & 0.91 \\ 
   6 & 0.364*** & 0.973*** & -0.007 & 0.023*** &  &  &  &  & -2230.10 & 0.90 \\ 
   7 & 0.364*** & 0.972*** &  &  &  &  &  &  & -2280.08 & 0.90 \\ 
   8 &  &  & -0.025 & 0.011 &  &  &  &  & -22270.55 & -0.00 \\ 
   9 & 0.286*** &  &  &  &  &  &  &  & -18986.63 & 0.32 \\ 
   10 &  & 0.798*** &  &  &  &  &  &  & -17714.99 & 0.41 \\ 
   11 &  &  & -0.008 &  &  &  &  &  & -22269.34 & -0.00 \\ 
   12 &  &  &  & 0.002 &  &  &  &  & -22269.51 & -0.00 \\ 
      \hline
   \end{tabular}
\caption{Elaboration and innovation as predictors of beak shape distance from centroid.
The table shows parameter estimates of predictors from 12 different phylogenetic generalised least squares (PGLS) models with $\Delta$AIC showing relative model fit following removal of terms. Models 1-5 are multiple regressions including all variables and one or more interaction terms (e.g. model 1 contains all the terms and their interactions); models 6-8 are are multiple regressions including two or more variables and no interaction terms (e.g. model 6 contains all the terms but no interactions); models 9-12 are single predictor models (e.g. model 9 contains only one term). Values in the table are based on PGLS in which Pagel’s $\lambda$ was fixed at 1 (i.e. assuming a Brownian motion model of evolution). Fixing $\lambda$ is necessary to allow model comparison with AIC. We also fitted the same models fixing $\lambda$ to 0.727 (the lowest value of $\lambda$ from any individual model) and relative model fit and interpretation of parameters was unaffected. $E$ = elaboration; $I$ = innovation; AIC = Akaike Information Criterion; *** = p-value $<$ 0.001.}
\label{Tab:pgls}
\end{table}
\end{landscape}
